--- 
title: "Statistical tests as linear models"
author: "Steve Doogue"
date: "`r Sys.Date()`"
bibliography:
- book.bib
- packages.bib
description: This is a minimal example of using the bookdown package to write a book.
  The output format for this example is bookdown::gitbook.
documentclass: book
link-citations: yes
site: bookdown::bookdown_site
biblio-style: apalike
---

# Introduction

This is a work through of the book [Common statistical tests are linear models (or: how to teach stats)](https://lindeloev.github.io/tests-as-linear/) by Jonas Kristoffer Lindelov. The book demonstrates how most common statistical models (t-test, correlation, ANOVA; chi-square, etc.) are special cases of linear models or a very close approximations.

This aims to solidy my understanding of the original book by replicating the examples providing, and by adding additional background information to fill gaps in my knowledge.

This is a _sample_ book written in **Markdown**. You can use anything that Pandoc's Markdown supports, e.g., a math equation $a^2 + b^2 = c^2$.

The **bookdown** package can be installed from CRAN or Github:

```{r eval=FALSE}
install.packages("bookdown")
# or the development version
# devtools::install_github("rstudio/bookdown")
```

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading `#`.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): <https://yihui.name/tinytex/>.

```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')
```
